### 2025-07-18 paper/quarto_aaai/paper.tex
Cut down to 7 pages

To make the desiderata for our framework more concrete, we follow the
previous work in tying the explanatory capacity of models to the quality
of CEs that can be generated for them
\citep{altmeyer2024faithful, augustin2020adversarial}. For simplicity,
we refer to ``explanatory capacity'' as ``explainability'' in the rest
of this manuscript (see Def. \ref{def-explainability} below for a more
formal exposition).

@@

It helps to look at these findings through the lens of representation
learning with high degrees of freedom. Deep ensembles are approximate
Bayesian model averages, which are particularly effective when models
are underspecified by the available data \citep{wilson2020case}.
Averaging across solutions mitigates the aforementioned risk of
overrelying on a single locally optimal representation that corresponds
to semantically meaningless explanations for the data. Likewise,
previous work of \citet{schut2021generating} found that generating
plausible (``interpretable'') CEs is almost trivial for deep ensembles
that have undergone adversarial training. The case for JEMs is even
clearer: they optimize a hybrid objective that induces both high
predictive performance and strong generative capacity
\citep{grathwohl2020your}, which bears resemblance to the idea of
aligning models with plausible explanations and has inspired our CT
objective.

@@

Counterfactual explanations are typically generated by solving
variations of the following optimization problem,
\begin{equation}\phantomsection\label{eq-general}{
\begin{aligned}
\min_{\mathbf{X}^\prime \in \mathcal{X}^D} \left\{  {\text{yloss}(\mathbf{M}_\theta(\mathbf{x}^{\prime}),\mathbf{y}^+)}+ \lambda {\text{reg}(\mathbf{x}^{\prime}) }  \right\} 
\end{aligned}
}\end{equation} where
\(\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denotes a
classifier, \(\mathbf{x}^{\prime}\) denotes the counterfactual with
\(D\) features and \(\mathbf{y}^+\) denotes some target class. The
\(\text{yloss}(\cdot)\) function quantifies the discrepancy between
current model predictions for \(\mathbf{x}^{\prime}\) and the target
class (a conventional choice is cross-entropy). Finally, we use
\(\text{reg}(\cdot)\) to denote any form of regularization used to
induce certain properties on the counterfactual. In their seminal paper,
\citet{wachter2017counterfactual} propose regularizing the distance
between counterfactuals and their original factual values, to ensure
that individuals seeking recourse through CE face minimal costs in terms
of feature changes. Different variations of Equation~\ref{eq-general}
have been proposed in the literature to address various other desiderata
including those already introduced above (faithfulness, plausibility and
actionability). Like \citet{wachter2017counterfactual}, most of these
approaches rely on gradient descent to optimize
Equation~\ref{eq-general}. For more details on various approaches tested
in this work, we refer the reader to the supplementary appendix. In the
following, we describe in detail how counterfactuals are generated and
used in CT.

@@

The principal goal of CT is to improve model explainability by aligning
models with faithful explanations that are plausible and actionable.
Formally, we define explainability as follows:

@@

\noindent The characterization of faithfulness and plausibility in Def.
\ref{def-explainability} is the same as in \citet{altmeyer2024faithful},
with adapted notation. Intuitively, plausible counterfactuals are
consistent with the data and faithful counterfactuals are consistent
with what the model has learned about the input data. Actionability
constraints in Def. \ref{def-explainability} vary and depend on the
context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
choose to only consider domain and mutability constraints for individual
features \(x_d\) for \(d=1,...,D\). We also limit ourselves to
classification tasks for reasons discussed in
Section~\ref{sec-discussion}.

@@

Test accuracies on adversarially perturbed data are shown in
Figure~\ref{fig-acc}. The perturbations size, \(\varepsilon\in[0,0.1]\),
increases along the horizontal axis and includes zero, corresponding to
standard test accuracy for non-perturbed data. For all synthetic
datasets, predictive performance of CT is virtually identical to the
baseline and unaffected by perturbations. For all real-world datasets,
we find that CT substantially improves robustness: while in some cases
baseline accuracies drop to essentially zero for large enough
perturbation sizes, accuracies of CT models are remarkably robust to
gradient-based adversarial attacks.

@@

For example, \emph{REVISE} tends to perform the worst, most likely
because it uses a surrogate VAE to generate counterfactuals which
impedes faithfulness \citep{altmeyer2024faithful}. Increasing \(T\), the
maximum number of steps, generally yields better outcomes because more
CEs can mature in each training epoch. The impact of \(\tau\), the
required decision threshold is more difficult to predict. On ``harder''
datasets it may be difficult to satisfy high \(\tau\) for any given
sample (i.e., also factuals) and so increasing this threshold does not
seem to correlate with better outcomes. In fact, \(\tau=0.5\) generally
leads to optimal results as it is associated with high proportions of
mature counterfactuals.

>>>>

To make the desiderata for our framework more concrete, we follow the
previous work in tying the explanatory capacity of models to the quality
of CEs that can be generated for them
\citep{altmeyer2024faithful, augustin2020adversarial}. For simplicity,
we refer to ``explanatory capacity'' as ``explainability'' in the rest
of this manuscript (see Def. \ref{def-explainability}).

@@

It helps to look at these findings through the lens of representation
learning with high degrees of freedom. Deep ensembles are approximate
Bayesian model averages, which are particularly effective when models
are underspecified by the available data \citep{wilson2020case}.
Averaging across solutions mitigates the aforementioned risk of
overrelying on a single locally optimal representation that corresponds
to semantically meaningless explanations. Likewise, previous work of
\citet{schut2021generating} found that generating plausible
(``interpretable'') CEs is almost trivial for deep ensembles that have
undergone adversarial training. The case for JEMs is even clearer: they
optimize a hybrid objective that induces both high predictive
performance and strong generative capacity \citep{grathwohl2020your},
which resembles the idea of aligning models with plausible explanations
and has inspired CT.

@@

Counterfactual explanations are typically generated by solving
variations of the following optimization problem,
\begin{equation}\phantomsection\label{eq-general}{
\begin{aligned}
\min_{\mathbf{X}^\prime \in \mathcal{X}^D} \left\{  {\text{yloss}(\mathbf{M}_\theta(\mathbf{x}^{\prime}),\mathbf{y}^+)}+ \lambda {\text{reg}(\mathbf{x}^{\prime}) }  \right\} 
\end{aligned}
}\end{equation} where
\(\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denotes a
classifier, \(\mathbf{x}^{\prime}\) denotes the counterfactual with
\(D\) features and \(\mathbf{y}^+\) denotes some target class. The
\(\text{yloss}(\cdot)\) function quantifies the discrepancy between
current model predictions for \(\mathbf{x}^{\prime}\) and the target
class (a conventional choice is cross-entropy). Finally, we use
\(\text{reg}(\cdot)\) to denote any form of regularization used to
induce certain properties on the counterfactual. In their seminal paper,
\citet{wachter2017counterfactual} propose regularizing the distance
between counterfactuals and their original factual values, to ensure
that individuals seeking recourse through CE face minimal costs in terms
of feature changes. Different variations of Equation~\ref{eq-general}
have been proposed in the literature to address various other desiderata
including those introduced above (faithfulness, plausibility and
actionability). Like \citet{wachter2017counterfactual}, most of these
approaches rely on gradient descent to optimize
Equation~\ref{eq-general}. For more details on various approaches tested
in this work, we refer the reader to the supplementary appendix. In the
following, we describe in detail how counterfactuals are generated and
used in CT.

@@

The goal of CT is to improve model explainability by aligning models
with faithful explanations that are plausible and actionable. Formally,
we define explainability as follows:

@@

The characterization of faithfulness and plausibility in Def.
\ref{def-explainability} is the same as in \citet{altmeyer2024faithful},
with adapted notation. Intuitively, plausible counterfactuals are
consistent with the data and faithful counterfactuals are consistent
with what the model has learned about the input data. Actionability
constraints in Def. \ref{def-explainability} vary and depend on the
context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
choose to only consider domain and mutability constraints for individual
features \(x_d\) for \(d=1,...,D\). We also limit ourselves to
classification tasks for reasons discussed in
Section~\ref{sec-discussion}.

@@

Test accuracies on adversarially perturbed data are shown in
Figure~\ref{fig-acc}. The perturbations size, \(\varepsilon\in[0,0.1]\),
increases along the horizontal axis and includes zero, corresponding to
standard test accuracy for non-perturbed data. For all synthetic
datasets, predictive performance of CT is virtually identical to the
baseline and unaffected by perturbations. For all real-world datasets,
we find that CT substantially improves robustness: while in some cases
baseline accuracies drop to essentially zero for large enough
perturbation sizes, accuracies of CT models remain remarkably robust.

@@

First, we look at hyperparameters that guide the gradient-based
counterfactual search. Increasing \(T\), the maximum number of steps,
generally yields better outcomes because more CEs can mature in each
training epoch. The impact of \(\tau\), the required decision threshold
is more difficult to predict. On ``harder'' datasets it may be difficult
to satisfy high \(\tau\) for any given sample (i.e., also factuals) and
so increasing this threshold does not seem to correlate with better
outcomes. In fact, \(\tau=0.5\) generally leads to optimal results as it
is associated with high proportions of mature counterfactuals.

### 2025-07-18 paper/sections/experiments.qmd
Cut down to 7 pages

Test accuracies on adversarially perturbed data are shown in @fig-acc. The perturbations size, $\varepsilon\in[0,0.1]$, increases along the horizontal axis and includes zero, corresponding to standard test accuracy for non-perturbed data. For all synthetic datasets, predictive performance of CT is virtually identical to the baseline and unaffected by perturbations. For all real-world datasets, we find that CT substantially improves robustness: while in some cases baseline accuracies drop to essentially zero for large enough perturbation sizes, accuracies of CT models are remarkably robust to gradient-based adversarial attacks.

@@

For example, *REVISE* tends to perform the worst, most likely because it uses a surrogate VAE to generate counterfactuals which impedes faithfulness [@altmeyer2024faithful]. Increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature in each training epoch. The impact of $\tau$, the required decision threshold is more difficult to predict. On "harder" datasets it may be difficult to satisfy high $\tau$ for any given sample (i.e., also factuals) and so increasing this threshold does not seem to correlate with better outcomes. In fact, $\tau=0.5$ generally leads to optimal results as it is associated with high proportions of mature counterfactuals.

>>>>

Test accuracies on adversarially perturbed data are shown in @fig-acc. The perturbations size, $\varepsilon\in[0,0.1]$, increases along the horizontal axis and includes zero, corresponding to standard test accuracy for non-perturbed data. For all synthetic datasets, predictive performance of CT is virtually identical to the baseline and unaffected by perturbations. For all real-world datasets, we find that CT substantially improves robustness: while in some cases baseline accuracies drop to essentially zero for large enough perturbation sizes, accuracies of CT models remain remarkably robust.

@@

First, we look at hyperparameters that guide the gradient-based counterfactual search. Increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature in each training epoch. The impact of $\tau$, the required decision threshold is more difficult to predict. On "harder" datasets it may be difficult to satisfy high $\tau$ for any given sample (i.e., also factuals) and so increasing this threshold does not seem to correlate with better outcomes. In fact, $\tau=0.5$ generally leads to optimal results as it is associated with high proportions of mature counterfactuals.

### 2025-07-18 paper/sections/method.qmd
Cut down to 7 pages

Counterfactual explanations are typically generated by solving variations of the following optimization problem,
$$
\begin{aligned}
\min_{\mathbf{X}^\prime \in \mathcal{X}^D} \left\{  {\text{yloss}(\mathbf{M}_\theta(\mathbf{x}^{\prime}),\mathbf{y}^+)}+ \lambda {\text{reg}(\mathbf{x}^{\prime}) }  \right\} 
\end{aligned}
$$ {#eq-general}
where $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denotes a classifier, $\mathbf{x}^{\prime}$ denotes the counterfactual with $D$ features and $\mathbf{y}^+$ denotes some target class. The $\text{yloss}(\cdot)$ function quantifies the discrepancy between current model predictions for $\mathbf{x}^{\prime}$ and the target class (a conventional choice is cross-entropy). Finally, we use $\text{reg}(\cdot)$ to denote any form of regularization used to induce certain properties on the counterfactual. In their seminal paper, @wachter2017counterfactual propose regularizing the distance between counterfactuals and their original factual values, to ensure that individuals seeking recourse through CE face minimal costs in terms of feature changes. Different variations of @eq-general have been proposed in the literature to address various other desiderata including those already introduced above (faithfulness, plausibility and actionability). Like @wachter2017counterfactual, most of these approaches rely on gradient descent to optimize @eq-general. For more details on various approaches tested in this work, we refer the reader to the supplementary appendix. In the following, we describe in detail how counterfactuals are generated and used in CT.

@@

`\noindent`{=latex} The characterization of faithfulness and plausibility in Def. \ref{def-explainability} is the same as in @altmeyer2024faithful, with adapted notation. Intuitively, plausible counterfactuals are consistent with the data and faithful counterfactuals are consistent with what the model has learned about the input data. Actionability constraints in Def. \ref{def-explainability} vary and depend on the context in which $\mathbf{M}_\theta$ is deployed. In this work, we choose to only consider domain and mutability constraints for individual features $x_d$ for $d=1,...,D$. We also limit ourselves to classification tasks for reasons discussed in @sec-discussion.

>>>>

Counterfactual explanations are typically generated by solving variations of the following optimization problem,
$$
\begin{aligned}
\min_{\mathbf{X}^\prime \in \mathcal{X}^D} \left\{  {\text{yloss}(\mathbf{M}_\theta(\mathbf{x}^{\prime}),\mathbf{y}^+)}+ \lambda {\text{reg}(\mathbf{x}^{\prime}) }  \right\} 
\end{aligned}
$$ {#eq-general}
where $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denotes a classifier, $\mathbf{x}^{\prime}$ denotes the counterfactual with $D$ features and $\mathbf{y}^+$ denotes some target class. The $\text{yloss}(\cdot)$ function quantifies the discrepancy between current model predictions for $\mathbf{x}^{\prime}$ and the target class (a conventional choice is cross-entropy). Finally, we use $\text{reg}(\cdot)$ to denote any form of regularization used to induce certain properties on the counterfactual. In their seminal paper, @wachter2017counterfactual propose regularizing the distance between counterfactuals and their original factual values, to ensure that individuals seeking recourse through CE face minimal costs in terms of feature changes. Different variations of @eq-general have been proposed in the literature to address various other desiderata including those introduced above (faithfulness, plausibility and actionability). Like @wachter2017counterfactual, most of these approaches rely on gradient descent to optimize @eq-general. For more details on various approaches tested in this work, we refer the reader to the supplementary appendix. In the following, we describe in detail how counterfactuals are generated and used in CT.

@@

The characterization of faithfulness and plausibility in Def. \ref{def-explainability} is the same as in @altmeyer2024faithful, with adapted notation. Intuitively, plausible counterfactuals are consistent with the data and faithful counterfactuals are consistent with what the model has learned about the input data. Actionability constraints in Def. \ref{def-explainability} vary and depend on the context in which $\mathbf{M}_\theta$ is deployed. In this work, we choose to only consider domain and mutability constraints for individual features $x_d$ for $d=1,...,D$. We also limit ourselves to classification tasks for reasons discussed in @sec-discussion.

### 2025-07-18 paper/sections/related.qmd
Cut down to 7 pages

To make the desiderata for our framework more concrete, we follow the previous work in tying the explanatory capacity of models to the quality of CEs that can be generated for them [@altmeyer2024faithful; @augustin2020adversarial]. For simplicity, we refer to "explanatory capacity" as "explainability" in the rest of this manuscript (see Def. \ref{def-explainability} below for a more formal exposition).

@@

It helps to look at these findings through the lens of representation learning with high degrees of freedom. Deep ensembles are approximate Bayesian model averages, which are particularly effective when models are underspecified by the available data [@wilson2020case]. Averaging across solutions mitigates the aforementioned risk of overrelying on a single locally optimal representation that corresponds to semantically meaningless explanations for the data. Likewise, previous work of @schut2021generating found that generating plausible ("interpretable") CEs is almost trivial for deep ensembles that have undergone adversarial training. The case for JEMs is even clearer: they optimize a hybrid objective that induces both high predictive performance and strong generative capacity [@grathwohl2020your], which bears resemblance to the idea of aligning models with plausible explanations and has inspired our CT objective.

>>>>

To make the desiderata for our framework more concrete, we follow the previous work in tying the explanatory capacity of models to the quality of CEs that can be generated for them [@altmeyer2024faithful; @augustin2020adversarial]. For simplicity, we refer to "explanatory capacity" as "explainability" in the rest of this manuscript (see Def. \ref{def-explainability}).

@@

It helps to look at these findings through the lens of representation learning with high degrees of freedom. Deep ensembles are approximate Bayesian model averages, which are particularly effective when models are underspecified by the available data [@wilson2020case]. Averaging across solutions mitigates the aforementioned risk of overrelying on a single locally optimal representation that corresponds to semantically meaningless explanations. Likewise, previous work of @schut2021generating found that generating plausible ("interpretable") CEs is almost trivial for deep ensembles that have undergone adversarial training. The case for JEMs is even clearer: they optimize a hybrid objective that induces both high predictive performance and strong generative capacity [@grathwohl2020your], which resembles the idea of aligning models with plausible explanations and has inspired CT.
